How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-21 Paper/paper.tex
corrected number
input on User interactivity in RQ2

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

		Projects utilizing variables & 73,577 & 31.51\% \\
		Projects utilizing lists & 9,358 & 4.01\% \\
		Projects with conditional statements & 92,959 & 39.81\% \\
		Projects with recursive loop statements & 180,210 & 77.18\% \\
		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		Projects with cloned scripts (>=5 blocks) across sprites & 59,634 & 25.54\% \\
		Projects with cloned scripts (>=5 blocks) within sprites & 23,671 & 10.14\%\\
		Projects with cloned procedures (>=5 blocks) & 4,945 & 2.12\%\\
		Projects with cloned functionality blocks (>=4 blocks) across sprites & 60,554 & 25.93\%\\
		Projects with exact clones (>=5 blocks) across sprites & 27,574 & 11.81\%\\
		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
		Projects containing dead code & 56,890	& 24.36\%\\
		Projects containing large scripts & 69,521 & 29.77\%\\
		Projects containing large sprites & 31,954 & 13.68\%\\		
		\hline
	\end{tabular}
	\caption{Characteristics in the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

@@

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
	\end{minipage}
}

@@

\subsection{Code Smells}
Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}. His book gives an overview of code smells and corresponding refactorings. Fowler's work was followed by efforts focused on the automatic identification of code smells by means of metrics. Marinescu~\cite{marinescu_detecting_2001} for instance, uses metrics to identify \emph{suspect} classes: classes which could have design flaws. Lanza and Marinescu~\cite{lanza_object-oriented_2006} explain this methodology in more detail. Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark. Olbrich \emph{et al.} furthermore investigates the changes in smells over time, and discusses their impact~\cite{olbrich_evolution_2009}. Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications. The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our \dup smell.

\subsection{Smells beyond the OO paradigm}
In recent times, code smells have been applied to programs outside of the regular programming domain. In our past work, we have, for example, studied code smells within spreadsheets, both at the formula level~\cite{hermans_detecting_2014} and between worksheets~\cite{hermans_detecting_2012}.

More recently, we compared two datasets: one containing spreadsheets which users found unmaintainable, and a version of the same spreadsheets rebuilt by professional spreadsheet developers. The results show that the improved versions suffered from smells to a lesser degree, increasing our confidence that presence of smells indeed coincides with users finding spreadsheets hard to maintain~\cite{jansen_code_2015}.

In addition to spreadsheets, code smells have also been studied in the context of Yahoo! Pipes a web mashup tool. An experiment demonstrated that users preferred the non-smelly versions of Yahoo Pipes programs~\cite{stolee_refactoring_2011}.

%\subsection{Agile practices in education}
%A third related line of work are research directions where other methods from agile development, like testing and refactoring, have been applied in education. For example, the use of test-driven development has been studied within the context of programming education at university level, using text-based languages in several studies. Most controlled experiments resulted in increased code quality for the group of students using TDD, for example [13] reported on a 35\% reduction in defects, while [14] found 45\%. For an extensive overview of TDD approaches in university programming courses, see [15]. \todo{referenties fixen}

\subsection{Quality of Scratch programs}
Finally, there are other works on the quality of Scratch programs. There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication. An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.

@@

>>>>

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		User input blocks per project&4.77&0&0&1&4&1889\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

		Projects utilizing variables & 73,577 & 31.51\% \\
		Projects utilizing lists & 9,358 & 4.01\% \\
		Projects with conditional statements & 92,959 & 39.81\% \\
		Projects accepting user input & 131,314 & 56.24\% \\
		Projects with recursive loop statements & 180,210 & 77.18\% \\
		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
		Projects with cloned scripts (>=5 blocks) across sprites & 59,634 & 25.54\% \\
		Projects with cloned scripts (>=5 blocks) within sprites & 23,671 & 10.14\%\\
		Projects with cloned procedures (>=5 blocks) & 4,945 & 2.12\%\\
		Projects with cloned functionality blocks (>=4 blocks) across sprites & 60,554 & 25.93\%\\
		Projects with exact clones (>=5 blocks) across sprites & 27,574 & 11.81\%\\
		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
		Projects containing dead code & 56,890	& 24.36\%\\
		Projects containing large scripts & 69,521 & 29.77\%\\
		Projects containing large sprites & 31,954 & 13.68\%\\		
		\hline
	\end{tabular}
	\caption{Characteristics in the projects in the dataset}
	\label{tbl-characteristics}
\end{table*}

@@

Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project. Table \ref{tbl-userinput} lists the frequency of use of user input controls. The most commonly used one is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects. The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.

\begin{table}
	\centering
	\begin{tabular}{lrr}
		\hline
		\textbf{Block}&\textbf{Projects}&\textbf{Occurrences}\\
		\hline
	\texttt{when <> key pressed}&71,096&294,771\\
	\texttt{when this sprite clicked}&39,179&198,342\\
	(Sensing) \texttt{key <> pressed?}&37,919&291,657\\
	(Sensing) \texttt{ask <> and wait}&19,039&66,850\\
	(Sensing) \texttt{mouse down?}&9,115&54,079\\
	(Sensing) \texttt{<attrib> of <>}&9,068&155,468\\
	(Sensing) \texttt{mouse X}&5,977&27,321\\
	(Sensing) \texttt{mouse Y}&3,940&22,035\\
	\texttt{when <sensor> > <value>}&705&1,570\\
	(Sensing) \texttt{video <> on <>}&434&1,397\\
		\hline
	\end{tabular}
	\caption{Frequency of use of user input blocks in the \nScriptPrograms projects of the dataset}
	\label{tbl-userinput}
\end{table}

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. More than half of the projects are interactive. 29.57\% of the projects utilize broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
	\end{minipage}
}

@@

The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education has received significant research attention during the past years. A number of studies have been carried out on the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.

Finally, there are other works on the quality of Scratch programs. There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication. An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.

@@

Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}. His book gives an overview of code smells and corresponding refactorings. Fowler's work was followed by efforts focused on the automatic identification of code smells by means of metrics. Marinescu~\cite{marinescu_detecting_2001} for instance, uses metrics to identify \emph{suspect} classes: classes which could have design flaws. Lanza and Marinescu~\cite{lanza_object-oriented_2006} explain this methodology in more detail. Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark. Olbrich \emph{et al.} furthermore investigates the changes in smells over time, and discusses their impact~\cite{olbrich_evolution_2009}. Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications. The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our \dup smell.
